\pkg{NeuralNetTools}: Visualization and Analysis Tools for Neural Networks

### 2015-11-30 nnt_manu.Rnw
manual bib formatting
revisions done from Marc's comments

%% an abstract and keywords
\Abstract{
Supervised neural networks have been applied as a machine learning technique to identify and predict emergent patterns in unstructured datasets.  A common criticism of the method is the inability to characterize relationships among variables from a fitted model. Although several methods have been proposed to `illuminate the black-box', they have not been made available in an open-source programming environment.  This article describes the \pkg{NeuralNetTools} package that can be used for the interpretation of supervised neural network models created in \proglang{R}.  Functions in the package can be used to visualize a model using a neural network interpretation diagram, evaluate variable importance by disaggregating the model weights, and perform a sensitivity analysis of the response variables to changes in the input variables.  Methods are provided for objects from many of the common neural networks packages in \proglang{R}, including \pkg{caret}, \pkg{neuralnet}, \pkg{nnet}, and \pkg{RSNNS}.  The article provides a brief overview of the theoretical foundation of neural networks, a description of the package structure and functions, and an applied example to provide a context for model development with \pkg{NeuralNetTools}.  Overall, the package provides a toolset for neural networks that complements existing quantitative techniques for data-intensive exploration.      
}
\Keywords{neural networks, plotnet, sensitivity, variable importance, \proglang{R}}
\Plainkeywords{neural networks, plotnet, sensitivity, variable importance, R} %% without formatting
%% at least one keyword must be supplied

@@

Data-intensive analysis is a relatively new research approach that focuses on the synthesis of unstructured information to identify patterns or trends `born from the data' \citep{Kelling09}.  A central theme is the focus on data exploration and prediction as compared to hypothesis-testing using domain-specific methods for scientific exploration \citep{Kell03}.  Demand for quantitative toolsets to address challenges in data-rich environments has increased drastically with the advancement of techniques for rapid acquisition of data. Fields of research characterized by high-throughput data have a strong foundation in computationally-intensive methods of analysis (e.g., \citet{Saeys07}), whereas disciplines that have historically been limited by data quantity have also realized the importance of quantitative toolsets given the development of novel techniques to acquire information (e.g., \citet{Swanson15}).  Quantitative methods that facilitate inductive reasoning can serve a complementary role to conventional, hypothesis-driven approaches to scientific discovery \citep{Kell03}.

Statistical methods that have been used to support data exploration are numerous \citep{Jain00}.  A common theme among data intensive methods is the use of machine-learning algorithms where the primary objective is to identify emergent patterns with minimal human intervention.  Neural networks, in particular, are designed to mimic the neuronal structure of the human brain by `learning' inherent data structures through adaptive algorithms \citep{Rumelhart86,Ripley96}.  Although the conceptual model was introduced several decades ago \citep{McCulloch43}, neural networks have had a central role in emerging fields focused on data exploration.  The most popular form of neural network is the feed-forward \ac{mlp} trained using the backpropagation algorithm \citep{Rumelhart86}.  This model is typically used to predict the response of one or more variables given one to many explanatory variables.  The hallmark feature of the \ac{mlp} is the characterization of relationships using an arbitrary number of parameters (i.e., the hidden layer) that are chosen through iterative training with the backpropation algorithm.  Conceptually, the \ac{mlp} is a hyper-parameterized non-linear model that can fit a smooth function to any dataset with minimal residual error \citep{Hornik91}.

An arbitrarily large number of parameters to fit a neural network provides obvious predictive advantages, but conversely complicates the extraction of model information.  Diagnostic information such as variable importance or model sensitivity are necessary aspects of exploratory data analysis that are not easily obtained from a neural network. As such, a common criticism is that neural networks are `black-boxes' that offer minimal insight into relationships among variables \citep[e.g.,][]{Paruelo97}.  \citet{Olden02} provide a rebuttal to this concern by describing methods to extract information from neural networks.  Many of these methods were previously described but not commonly used.  For example, \citet{Olden02} describe  the \ac{nid} for plotting \citep{Ozesmi99}, the Garson algorithm for variable importance \citep{Garson91}, and the Profile method for sensitivity analysis \citep{Lek96}.  These quantitative tools `illuminate the black box' by disaggregating the network parameters to characterize relationships between variables that are described by the model.  Although \ac{mlp} neural networks were developed for prediction, methods described in \citet{Olden02} leverage these models to describe data signals.  Increasing the accesibility of these diagnostic tools will have value for exploratory data analysis and may also inform causal inference.

This article describes the \pkg{NeuralNetTools} package for \proglang{R} that was developed to improve the breadth and quality of information obtained from the \ac{mlp} neural network.  Functions provided by the package are those described in \citet{Olden02} but have not been previously available in an open-source programming environment.  The reach of the package is extensive such that generic functions were developed for model objects from the most popular neural network packages available in \proglang{R}.  The objectives are to 1) provide an overview of the statistical foundation of the \ac{mlp} network, 2) describe the theory and application of the main functions in the \pkg{NeuralNetTools} package, and 3) provide an applied example using neural networks and \pkg{NeuralNetTools} in data exploration.  The package is currently available on the \ac{CRAN}, whereas the development version is maintained as a GitHub repository.

@@

The typical \ac{mlp} network is composed of multiple layers that define the transfer of information between input and response layers.  Information travels in one direction where a set of values for variables in the input layer propagates through one or more hidden layers to the final layer of the response variables. `Hidden' layers between the input and response layers are key components of a neural network that mediate the transfer of information.  Just as the input and response layers are composed of variables or `nodes', each hidden layer is composed of nodes with weighted connections that define the strength of information flow between layers.  `Bias' layers connected to hidden and response layers may also be used that are analagous to intercept terms in a standard regression model.

Training a neural network model requires identifying the `optimal' weights that define the connections between the model layers.  The optimal weights are those that minimize prediction error for a test dataset that is independent of the training dataset.  Training is commonly achieved using the backpropagation algorithm described in \citep{Rumelhart86}.  This algorithm identifies the optimal weighting scheme through an iterative process where weights are gradually changed through a forward- and backward-propagation process \citep{Rumelhart86,Lek00}.  The algorithm begins by assigning an arbitrary weighting scheme to the connections in the network, followed by estimating the output in the response variable through the forward-propagation of information through the network, and finally calculating the difference between the predicted and actual value of the response.  The weights are then changed through a back-propagation step that begins by changing weights in the output layer and then the remaining hidden layers.  The process is repeated until the chosen error function is minimized, as in standard model-fitting techniques for regression \citep{Cheng94}.  Formulaically, a fitted \ac{mlp} neural network can be represented as \citep{Bishop95,Venables02}:
\begin{equation}
y_k = f_o \left(\sum\limits_{h} w_{hk}f_h \left( \sum\limits_{i} w_{ih}x_i\right) \right)
\end{equation}
where the estimated value $k$ of the response variable $y$ is a summation of the products between the respective weights $w$ for $i$ input variables $x$ and $h$ hidden nodes, mediated by the activation functions $f_h$ and $f_o$ for each hidden and output node.

@@

Methods in \pkg{NeuralNetTools} were written for several \proglang{R} packages that can be used to create \ac{mlp} neural networks: \pkg{neuralnet} \citep{Fritsch12}, \pkg{nnet} \citep{Venables02}, and \pkg{RSNNS} \citep{Bergmeir12}. Limited methods were also developed for neural network objects created with the \code{train} function from the \pkg{caret} package \citep{Kuhn15}.  Additional \proglang{R} packages that can create \ac{mlp} neural networks include \pkg{AMORE} that implements the ``TAO-robust back-propagation algorithm'' for model fitting \citep{Castejon14}, \pkg{FCNN4R} as an \proglang{R} interface to the \pkg{FCNN} \proglang{C++} library \citep{Klima15}, \pkg{monmlp} for networks with partial monotonicity constraints \citep{Cannon15}, and \pkg{qrnn} for quantile regression neural networks \citep{Cannon11}.  At the time of writing, the \ac{CRAN} download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included \Sexpr{round(tots_nnt)}\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over \Sexpr{round(tots_nnet)}\%.  As such, methods have not been included in \pkg{NeuralNetTools} for the remaining packages, although further development of \pkg{NeuralNetTools} could include additional methods based on popularity.  Methods for each function are currently available for  \code{mlp} (\pkg{RSNNS}), \code{nn} (\pkg{neuralnet}), \code{nnet} (\pkg{nnet}), and \code{train} (\pkg{caret}, only if the object also inherits from the \code{nnet} class.  Additional \code{default} or \code{numeric} methods are available for some of the generic functions.

@@

\pkg{NeuralNetTools} includes four main functions that were developed following similar techniques in \citet{Olden02} and references therein.  Functions include \code{plotnet} to plot a neural network interpretation diagram, \code{garson} and \code{olden} functions to evaluate variable importance, and \code{lekprofile} for a sensitivity analysis of neural network response to input variables.  Most of the functions require the extraction of model weights in a common format for the neural network object classes in \proglang{R}.  The \code{neuralweights} function can be used to retrieve model weights for any of the model classes describe above.  A two-element \code{list} is returned with the first element describing the structure of the network (number of nodes in the input, hidden, and output layers) and the second element as a named list of model weights.  The function is used internally within the main functions but may be useful for comparing networks of different classes.

@@

Existing plot functions in \proglang{R} to view neural networks is minimal.  Such tools have practical use for visualizing network architecture and connections between layers that mediate variable importance. To our knowledge, only the \pkg{neuralnet} and \pkg{FCNN4R} packages provide plotting methods for \ac{mlp} networks in \proglang{R}.  Although useful for viewing the basic structure, the output is minimal and does not include extensive options for customization.

@@

<<plotprune, eval = T, echo = T, fig.show = 'hold', fig.height = 7, fig.width = 7.7, out.width = '0.48\\textwidth', fig.cap = 'A pruned neural network from \\pkg{RSNNS} \\citep{Bergmeir12} using the ``Optimal Brain Surgeon\'\' algorithm described in \\citet{Zell98}.  The default plotting behavior of \\code{plotnet} is to omit pruned connections (\\ref{fig:plotprune1}), whereas they can be viewed as dashed lines by including the \\code{prune\\_col} argument (\\ref{fig:plotprune2}).', fig.subcap = c('', '')>>=
# pruned model using RSNNS
pruneFuncParams <- list(max_pr_error_increase = 10.0, pr_accepted_error = 1.0,
  no_of_pr_retrain_cycles = 1000, min_error_to_stop = 0.01, 
  init_matrix_value = 1e-6, input_pruning = TRUE, hidden_pruning = TRUE)
mod <- mlp(x, y, size = 5, pruneFunc = "OptimalBrainSurgeon",
 pruneFuncParams = pruneFuncParams)

@@

Although \pkg{NeuralnetTools} provides several methods to extract information from a fitted neural network, it does not provide explicit guidance for developing the initial model.  A potentially more challenging aspect of using \ac{mlp} neural networks is understanding the effects of network architecture on model performance, appropriate use of training and validation datasets, and implications for the bias-variance tradeoff with model over- or under-fitting \citep{Maier00}.  A detailed discussion of these issues is beyond the scope of this paper, although an example application is presented below to emphasize the importance of these considerations.  The models presented above, including the \code{neuraldat} dataset, are contrived examples to illustrate use of the \pkg{NeuralNetTools} package and they do not demonstrate a comprehensive or practical application of model development.  In general, the following should be considered during initial development \cite{Ripley96, Lek00, Maier00}:
\begin{itemize}
\item Initial data pre-processing to normalize inputs, standardize response, and assess influence of outliers,
\item Network architecture including number of hidden layers, number of nodes in each hidden layer, inclusion of bias or skip layers, and pruning weights or inputs,
\item Separating data into training and test datasets, e.g., 2:1, 3:1, leave-one-out, etc., 
\item Initial starting weights for the back-propagation algorithm, and
\item Criteria for stopping model training, e.g., error convergance tolerance, maximum number of iterations, minimum error on test dataset, etc.
\end{itemize}

A dataset from \pkg{nycflights13} \citep{Wickham14b} is used to demonstrate effects of training conditions on model conclusions.  This dataset provides information on all flights departing New York City (i.e., JFK, LGA, or EWR) in 2013.  The example uses all flights from the `UA' carrier in the month of December to identify variables that potentially influence arrival delays (\code{arr\_delay}, minutes) at the destination airport.  Factors potentially related to delays were selected from the dataset and included departure delay (\code{dep\_delay}, minutes), departure time (\code{dep\_time}, hours, minutes), arrival time (\code{arr\_time}, hours, minutes), travel time between desinations (\code{air\_time}, minutes), and distance flown (\code{distance}, miles).  All explanatory variables were normalized and the response variable was standardized to 0--1.

The effects of network architecture and initial starting weights on estimates of variable importance were evaluated by creating many models and assessing the uncertainty of the importance.  Models with one, five, or ten hidden nodes and 100 separate models for each node level were created.  Each model had a random set of starting weights for the first training iteration.  Importance estimates using \code{olden} were saved for each model and combined in a single plot to show overall variable importance as the median and 5\textsuperscript{th}/95\textsuperscript{th} percentiles from the 100 models for each node level (get \href{https://raw.githubusercontent.com/fawda123/nnt_manu/master/flightimp.R}{code}).

@@

A more compelling issue is when and how to apply neural networks given alternative methods of analysis.  The popularity of the \ac{mlp} neural network is partially the cause of generalizations and misperceptions of their benefits as modelling tools \citep{Burke97}. Perhaps an overstatement, the `neural' component is commonly advertised as a mathematical representation of the network of synaptic impulses in the human brain.  Additionally, several examples have shown that the \ac{mlp} network may provide comparable predictive performance as similar statistical methods \citep{Feng02,Razi05,Beck14a}.  A neural network should be considered a tool in the larger toolbox of data-intensive methods that should be used after examination of the tradeoffs between techniques, with particular emphasis on the specific needs of a dataset or research question.  Considerations for choosing a method may include power given the sample size, expected linear or non-linear interactions between variables, distributional forms of the response, and other relevant considerations of exploratory data analysis \citep{Zuur10}.  \pkg{NeuralNetTools} provides analysis tools that can inform a more balanced assessment given alternative and complementary methods for data exploration.

@@

I thank Bruce Vondracek, Sanford Weisberg, and Bruce Wilson of the University of Minnesota for general guidance during the development of this package.  Thanks to Sehan Lee for reviewing an earlier draft.  Contributions and suggestions from online users have also greatly improved the utility of the package.  Funding for this project was supported in part by an Interdisciplinary Doctoral Fellowship provided by the Graduate School at the University of Minnesota to M. Beck.

% note that ref titles need to be in title-case 
\bibliographystyle{jss}
\bibliography{refs}

>>>>

%% an abstract and keywords
\Abstract{
Supervised neural networks have been applied as a machine learning technique to identify and predict emergent patterns in unstructured datasets.  A common criticism of the method is the inability to characterize relationships among variables from a fitted model. Although several methods have been proposed to `illuminate the black box', they have not been made available in an open-source programming environment.  This article describes the \pkg{NeuralNetTools} package that can be used for the interpretation of supervised neural network models created in \proglang{R}.  Functions in the package can be used to visualize a model using a neural network interpretation diagram, evaluate variable importance by disaggregating the model weights, and perform a sensitivity analysis of the response variables to changes in the input variables.  Methods are provided for objects from many of the common neural networks packages in \proglang{R}, including \pkg{caret}, \pkg{neuralnet}, \pkg{nnet}, and \pkg{RSNNS}.  The article provides a brief overview of the theoretical foundation of neural networks, a description of the package structure and functions, and an applied example to provide a context for model development with \pkg{NeuralNetTools}.  Overall, the package provides a toolset for neural networks that complements existing quantitative techniques for data-intensive exploration.      
}
\Keywords{neural networks, plotnet, sensitivity, variable importance, \proglang{R}}
\Plainkeywords{neural networks, plotnet, sensitivity, variable importance, R} %% without formatting
%% at least one keyword must be supplied

@@

Data-intensive analysis is a relatively new research approach that focuses on the synthesis of unstructured information to identify patterns or trends `born from the data' \citep{Bell09,Kelling09,Michener12}.  A central theme is the focus on data exploration and prediction as compared to hypothesis-testing using domain-specific methods for scientific exploration \citep{Kell03}.  Demand for quantitative toolsets to address challenges in data-rich environments has increased drastically with the advancement of techniques for rapid acquisition of data. Fields of research characterized by high-throughput data (e.g., bioinformatics, \citet{Saeys07}) have a strong foundation in computationally-intensive methods of analysis, whereas disciplines that have historically been limited by data quantity (e.g., field ecology, \citet{Swanson15}) have also realized the importance of quantitative toolsets given the development of novel techniques to acquire information.  Quantitative methods that facilitate inductive reasoning can serve a complementary role to conventional, hypothesis-driven approaches to scientific discovery \citep{Kell03}.

Statistical methods that have been used to support data exploration are numerous \citep{Jain00,Recknagel06,Zuur10}.  A common theme among data intensive methods is the use of machine-learning algorithms where the primary objective is to identify emergent patterns with minimal human intervention.  Neural networks, in particular, are designed to mimic the neuronal structure of the human brain by `learning' inherent data structures through adaptive algorithms \citep{Rumelhart86,Ripley96}.  Although the conceptual model was introduced several decades ago \citep{McCulloch43}, neural networks have had a central role in emerging fields focused on data exploration.  The most popular form of neural network is the feed-forward \ac{mlp} trained using the backpropagation algorithm \citep{Rumelhart86}.  This model is typically used to predict the response of one or more variables given one to many explanatory variables.  The hallmark feature of the \ac{mlp} is the characterization of relationships using an arbitrary number of parameters (i.e., the hidden layer) that are chosen through iterative training with the backpropagation algorithm.  Conceptually, the \ac{mlp} is a hyper-parameterized non-linear model that can fit a smooth function to any dataset with minimal residual error \citep{Hornik91}.

An arbitrarily large number of parameters to fit a neural network provides obvious predictive advantages, but conversely complicates the extraction of model information.  Diagnostic information such as variable importance or model sensitivity are necessary aspects of exploratory data analysis that are not easily obtained from a neural network. As such, a common criticism is that neural networks are `black boxes' that offer minimal insight into relationships among variables \citep[e.g.,][]{Paruelo97}.  \citet{Olden02} provide a rebuttal to this concern by describing methods to extract information from neural networks.  Many of these methods were previously described but not commonly used.  For example, \citet{Olden02} describe  the \ac{nid} for plotting \citep{Ozesmi99}, the Garson algorithm for variable importance \citep{Garson91}, and the Profile method for sensitivity analysis \citep{Lek96}.  These quantitative tools `illuminate the black box' by disaggregating the network parameters to characterize relationships between variables that are described by the model.  Although \ac{mlp} neural networks were developed for prediction, methods described in \citet{Olden02} leverage these models to describe data signals.  Increasing the accesibility of these diagnostic tools will have value for exploratory data analysis and may also inform causal inference.

This article describes the \pkg{NeuralNetTools} package for \proglang{R} that was developed to improve the breadth and quality of information obtained from the \ac{mlp} neural network.  Functions provided by the package are those described in \citet{Olden02} but have not been previously available in an open-source programming environment.  The reach of the package is extensive in that generic functions were developed for model objects from the most popular neural network packages available in \proglang{R}.  The objectives of this article are to 1) provide an overview of the statistical foundation of the \ac{mlp} network, 2) describe the theory and application of the main functions in the \pkg{NeuralNetTools} package, and 3) provide an applied example using neural networks and \pkg{NeuralNetTools} in data exploration.  The package is currently available on the \ac{CRAN}, whereas the development version is maintained as a GitHub repository.

@@

The typical \ac{mlp} network is composed of multiple layers that define the transfer of information between input and response layers.  Information travels in one direction where a set of values for variables in the input layer propagates through one or more hidden layers to the final layer of the response variables. Hidden layers between the input and response layers are key components of a neural network that mediate the transfer of information.  Just as the input and response layers are composed of variables or nodes, each hidden layer is composed of nodes with weighted connections that define the strength of information flow between layers.  Bias layers connected to hidden and response layers may also be used that are analagous to intercept terms in a standard regression model.

Training a neural network model requires identifying the optimal weights that define the connections between the model layers.  The optimal weights are those that minimize prediction error for a test dataset that is independent of the training dataset.  Training is commonly achieved using the backpropagation algorithm described in \citep{Rumelhart86}.  This algorithm identifies the optimal weighting scheme through an iterative process where weights are gradually changed through a forward- and backward-propagation process \citep{Rumelhart86,Lek00}.  The algorithm begins by assigning an arbitrary weighting scheme to the connections in the network, followed by estimating the output in the response variable through the forward-propagation of information through the network, and finally calculating the difference between the predicted and actual value of the response.  The weights are then changed through a back-propagation step that begins by changing weights in the output layer and then the remaining hidden layers.  The process is repeated until the chosen error function is minimized, as in standard model-fitting techniques for regression \citep{Cheng94}.  Formulaically, a fitted \ac{mlp} neural network can be represented as \citep{Bishop95,Venables02}:
\begin{equation}
y_k = f_o \left(\sum\limits_{h} w_{hk}f_h \left( \sum\limits_{i} w_{ih}x_i\right) \right)
\end{equation}
where the estimated value $k$ of the response variable $y$ is a summation of the products between the respective weights $w$ for $i$ input variables $x$ and $h$ hidden nodes, mediated by the activation functions $f_h$ and $f_o$ for each hidden and output node.

@@

Methods in \pkg{NeuralNetTools} were written for several \proglang{R} packages that can be used to create \ac{mlp} neural networks: \pkg{neuralnet} \citep{Fritsch12}, \pkg{nnet} \citep{Venables02}, and \pkg{RSNNS} \citep{Bergmeir12}. Limited methods were also developed for neural network objects created with the \code{train} function from the \pkg{caret} package \citep{Kuhn15}.  Additional \proglang{R} packages that can create \ac{mlp} neural networks include \pkg{AMORE} that implements the `TAO-robust back-propagation algorithm' for model fitting \citep{Castejon14}, \pkg{FCNN4R} as an \proglang{R} interface to the \pkg{FCNN} \proglang{C++} library \citep{Klima15}, \pkg{monmlp} for networks with partial monotonicity constraints \citep{Cannon15}, and \pkg{qrnn} for quantile regression neural networks \citep{Cannon11}.  At the time of writing, the \ac{CRAN} download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included \Sexpr{round(tots_nnt)}\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over \Sexpr{round(tots_nnet)}\%.  As such, methods have not been included in \pkg{NeuralNetTools} for the remaining packages, although further development of \pkg{NeuralNetTools} could include additional methods based on popularity.  Methods for each function are currently available for  \code{mlp} (\pkg{RSNNS}), \code{nn} (\pkg{neuralnet}), \code{nnet} (\pkg{nnet}), and \code{train} (\pkg{caret}, only if the object also inherits from the \code{nnet} class.  Additional \code{default} or \code{numeric} methods are available for some of the generic functions.

@@

\pkg{NeuralNetTools} includes four main functions that were developed following similar techniques in \citet{Olden02} and references therein.  Functions include \code{plotnet} to plot a neural network interpretation diagram, \code{garson} and \code{olden} functions to evaluate variable importance, and \code{lekprofile} for a sensitivity analysis of neural network response to input variables.  Most of the functions require the extraction of model weights in a common format for the neural network object classes in \proglang{R}.  The \code{neuralweights} function can be used to retrieve model weights for any of the model classes described above.  A two-element \code{list} is returned with the first element describing the structure of the network (number of nodes in the input, hidden, and output layers) and the second element as a named list of model weights.  The function is used internally within the main functions but may be useful for comparing networks of different classes.

@@

Existing plot functions in \proglang{R} to view neural networks are minimal.  Such tools have practical use for visualizing network architecture and connections between layers that mediate variable importance. To our knowledge, only the \pkg{neuralnet} and \pkg{FCNN4R} packages provide plotting methods for \ac{mlp} networks in \proglang{R}.  Although useful for viewing the basic structure, the output is minimal and does not include extensive options for customization.

@@

<<plotprune, eval = T, echo = T, fig.show = 'hold', fig.height = 7, fig.width = 7.7, out.width = '0.48\\textwidth', fig.cap = 'A pruned neural network from \\pkg{RSNNS} \\citep{Bergmeir12} using the `Optimal Brain Surgeon\' algorithm described in \\citet{Zell98}.  The default plotting behavior of \\code{plotnet} is to omit pruned connections (\\ref{fig:plotprune1}), whereas they can be viewed as dashed lines by including the \\code{prune\\_col} argument (\\ref{fig:plotprune2}).', fig.subcap = c('', '')>>=
# pruned model using RSNNS
pruneFuncParams <- list(max_pr_error_increase = 10.0, pr_accepted_error = 1.0,
  no_of_pr_retrain_cycles = 1000, min_error_to_stop = 0.01, 
  init_matrix_value = 1e-6, input_pruning = TRUE, hidden_pruning = TRUE)
mod <- mlp(x, y, size = 5, pruneFunc = "OptimalBrainSurgeon",
 pruneFuncParams = pruneFuncParams)

@@

Although \pkg{NeuralnetTools} provides several methods to extract information from a fitted neural network, it does not provide explicit guidance for developing the initial model.  A potentially more challenging aspect of using \ac{mlp} neural networks is understanding the effects of network architecture on model performance, appropriate use of training and validation datasets, and implications for the bias-variance tradeoff with model over- or under-fitting \citep{Maier00}.  A detailed discussion of these issues is beyond the scope of this paper, although an example application is presented below to emphasize the importance of these considerations.  The models presented above, including the \code{neuraldat} dataset, are contrived examples to illustrate use of the \pkg{NeuralNetTools} package and they do not demonstrate a comprehensive or practical application of model development.  In general, the following should be considered during initial development \citep{Ripley96, Lek00, Maier00}:
\begin{itemize}
\item Initial data pre-processing to normalize inputs, standardize response, and assess influence of outliers,
\item Network architecture including number of hidden layers, number of nodes in each hidden layer, inclusion of bias or skip layers, and pruning weights or inputs,
\item Separating data into training and test datasets, e.g., 2:1, 3:1, leave-one-out, etc., 
\item Initial starting weights for the back-propagation algorithm, and
\item Criteria for stopping model training, e.g., error convergance tolerance, maximum number of iterations, minimum error on test dataset, etc.
\end{itemize}

A dataset from \pkg{nycflights13} \citep{Wickham14b} is used to demonstrate effects of training conditions on model conclusions (get analysis code \href{https://raw.githubusercontent.com/fawda123/nnt_manu/master/flightimp.R}{here}).  This dataset provides information on all flights departing New York City (i.e., JFK, LGA, or EWR) in 2013.  The example uses all flights from the UA carrier in the month of December to identify variables that potentially influence arrival delays (\code{arr\_delay}, minutes) at the destination airport.  Factors potentially related to delays were selected from the dataset and included departure delay (\code{dep\_delay}, minutes), departure time (\code{dep\_time}, hours, minutes), arrival time (\code{arr\_time}, hours, minutes), travel time between desinations (\code{air\_time}, minutes), and distance flown (\code{distance}, miles).  All explanatory variables were normalized and the response variable was standardized to 0--1.

The effects of network architecture and initial starting weights on estimates of variable importance were evaluated by creating many models and assessing the uncertainty of the importance.  Models with one, five, or ten hidden nodes and 100 separate models for each node level were created.  Each model had a random set of starting weights for the first training iteration.  Importance estimates using \code{olden} were saved for each model and combined in a single plot to show overall variable importance as the median and 5\textsuperscript{th}/95\textsuperscript{th} percentiles from the 100 models for each node level (see analysis \href{https://raw.githubusercontent.com/fawda123/nnt_manu/master/flightimp.R}{code}).

@@

A more compelling issue is when and how to apply neural networks given alternative methods of analysis.  The popularity of the \ac{mlp} neural network is partially the cause of generalizations and misperceptions of their benefits as modelling tools \citep{Burke97}. Perhaps an overstatement, the neural component is commonly advertised as a mathematical representation of the network of synaptic impulses in the human brain.  Additionally, several examples have shown that the \ac{mlp} network may provide comparable predictive performance as similar statistical methods \citep{Feng02,Razi05,Beck14a}.  A neural network should be considered a tool in the larger toolbox of data-intensive methods that should be used after examination of the tradeoffs between techniques, with particular emphasis on the specific needs of a dataset or research question.  Considerations for choosing a method may include power given the sample size, expected linear or non-linear interactions between variables, distributional forms of the response, and other relevant considerations of exploratory data analysis \citep{Zuur10}.  \pkg{NeuralNetTools} provides analysis tools that can inform a more balanced assessment given alternative and complementary methods for data exploration.

@@

I thank Bruce Vondracek, Sanford Weisberg, and Bruce Wilson of the University of Minnesota for general guidance during the development of this package.  Thanks to Sehan Lee and Marc Weber for reviewing an earlier draft.  Contributions and suggestions from online users have also greatly improved the utility of the package.  Funding for this project was supported in part by an Interdisciplinary Doctoral Fellowship provided by the Graduate School at the University of Minnesota to M. Beck.

% note that ref titles need to be in title-case 
% \bibliographystyle{jss}
% \bibliography{refs}
% this file is bbl edited by hand after running bibtex
\input{CopyOfnnt_manu.bbl}
